Finiteness in Cubical Type Theory

### 2020-06-14 sections/introduction.tex
finished countdown in introduction

Our strategy for finding solutions to a given puzzle is to describe precisely
the type of solutions to a puzzle, and then show that that type is finite.
So what is a ``solution'' to

\section{Notation and Background}
\subsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into \(\mathbf{Type}\).
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}

>>>>

Our strategy for finding solutions to a given puzzle is to describe precisely
the type of solutions to a puzzle, and then show that that type is finite.
So what is a ``solution'' to a countdown puzzle?
Broadly, it has two parts:
\begin{description}
  \item[A Transformation] from a list of numbers to an expression.
  \item[A Predicate] showing that the expression is valid and evaluates to the
    target.
\end{description}
The first part is described in Figure~\ref{countdown-transform}.

\input{figures/countdown-transformation}

This transformation has four steps.
First (Fig.~\ref{countdown-selection}) we have to pick which numbers we include
in our solution.
The representation of this transformation should be fairly simple: to pick from
\(n\) objects we can have an \(n\)-tuple of booleans.

Secondly (Fig.~\ref{countdown-permutation}) we have to permute the chosen
numbers.
The representation for a permutation is a little trickier to envision: proving
that it's finite is trickier still.
We will need to rely on some of the more involved lemmas later on for this
problem.

The third step (Fig.~\ref{countdown-operators}) is quite like the first: it's
just another \(n\)-tuple of finite objects (in this case operators chosen from
\(+\), \(\times\), \(-\), and \(\div\)).
Although it is complicated slightly by the fact that the \(n\) in this
\(n\)-tuple is dependent on the number of trues in the previously chosen tuple
of booleans (in terms of types, that means we'll need a \(\Sigma\) rather than a
\(\times\), explanations of which are forthcoming).

Finally (Fig.~\ref{countdown-parens}), we have to parenthesise the expression in
a certain way.
This can be encapsulated by a binary tree with a certain number of leaves:
proving that that is finite is tricky again.

Once we have proven that there are finitely many transformations for a list of
numbers, we will then have to filter them down to those transformations which
are valid, and evaluate to the target.
This amounts to proving that the decidable subset of a finite set is also
finite.

Finally, we will also want to optimise our solutions and solver: for this we
will remove equivalent expressions, which can be accomplished with quotients.
\section{Notation and Background}
\subsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into \(\mathbf{Type}\).
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
  \item[Lambdas] \todo{Figure out how to describe lambdas}
\end{description}
\subsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}
